---
title: "Basic R"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

# Basic R

## Writing in R and Rmarkdown

### Chatting with R

Using R is just a chat with the computer.

"Hey, R. What is $1 + 2$?"

```{r}
1 + 2
```

### Rmarkdown tricks

- To make text **bold**, we add two \*\*s around it.
- To make text *italicized*, we add just one \* around it.
- If we need special characters (like \* or \$), then we just add a forward "\\" in front of them (but not behind).
- Math symbols in your text are process with Latex, just put an "\$" before and after your math. Like this, \$y = x\$ becomes $y = x$. 

- Code blocks

To make a code block, press CTRL+ALT+I. Remember you can change the output of a code block by modifying some of its options, like below.

```{r}
banana <- 5

banana + 1
```


## Variables

### The assignment operator

We make variables with the `<-` operator.

```{r}
my_special_var <- 1 + 2

my_special_var
```

You can TECHNICALLY use `=` for assignment too. Never do this.

```{r}
my_other_var = 12

my_other_var + my_special_var
```

### Numerics

**Doubles**

Doubles are decimal numbers, like $1.1, 2.2, 3.0$. If I make a number variable without doing anything special, R defaults to a double.

```{r}
a <- 1.1
b <- 2.0

is.double(a)
is.double(b)
```

**Integers**

Integers must have an `L` after them. That is how R knows that you don't want a double, but instead want a "long-capable integer".

```{r}
c <- 1L
d <- 1

is.integer(c)
is.integer(d)
```

Here is a useful cheatsheet for the different numeric operators and how they behave.
  
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

data.frame(
  a = rep(10, 7),
  b = rep(3, 7),
  operator = c('+', '-', '*', '/', '^', '%/%', '%%')
  ) %>%
  mutate(
    ex = paste(a, operator, b),
    result = ex %>%
      map(rlang::parse_expr) %>%
      map(eval) %>%
      unlist() %>%
      round(3) %>%
      formatC(digits = 3, format = 'f', drop0trailing = T)
  ) %>%
  select(-a, -b) %>%
  knitr::kable(
    col.names = c('Operator', 'Expression', 'Result'),
    align = 'c')
```
  
**Why care about the difference?**

Almost 99% of the time, this wont matter. But, with big data, integers take up must less memory.

```{r}
my_integers <- seq(from = 1L, to = 1e6L, by = 1L)
my_doubles <- seq(from = 1.0, to = 1e6, by = 1.0)

object.size(my_integers)
object.size(my_doubles)
```

Note here that although we are using only whole numbers from 1 to 1 million, the first sequence (`my_integers`) is stored as an integer and the second sequence (`my_doubles`) is stored as a number that may include decimals. This second case needs more space (twice as much) to be allocated in advance, even if we never use those decimal places.

Again, this will almost never matter for most people, most of the time. However, it is good to be aware of for when your datasets get large (i.e., several million cases or more).

### Characters

Characters are text symbols and they are made with either \"\" or \'\', either works.

```{r}
a <- 'here is someone\'s text'
b <- "here is more text"

a
b
```

To combine two strings, I use `paste()`.

```{r}
paste(a, b)
```

If I dont want a space, then I used `paste0()`.

```{r}
paste0(a, b)
```

### Booleans

These are True and False values. You make them with the symbols `T` or `TRUE` and `F` or `FALSE`.

```{r}
x <- T
y <- F
```

To compare them, we can use three operators.

- `&` is "and"
- `|` is "or"
- `!` is "not" (just give me the opposite of whatever is after me)

```{r}
x & y # false
x | y # true
x & !y # true
```

We can also have nested equations

```{r}
z <- F

x & !(y | z) # true
```

We can also compare numbers.

```{r}
a <- 1
b <- 2
```

```{r}
a < 1
a <= 1
a == 1
```

If I want to compare multiple numbers, I need to do it seperately.

```{r}
(a > 1) | (b > 1)
```

Remember that booleans are ultimately numeric values underneath.

```{r}
d <- T
k <- F
u <- 5

d*u
d*k

as.numeric(d)
as.numeric(k)
```


### Special types

`NA` - missing
    
```{r}
is.na(NA)
```

`NaN` - you did math wrong

```{r}
0/0
```

`Inf` - infinity

```{r}
-5/0
```


## Vectors

R is built is on vectors. Vectors are collections of a bunch of values of the same type.

```{r}
my_vec <- c(1, 5, 3, 7)

my_vec
```

If I try to put different types together, they go to the most primitive type (usually a character string).

```{r}
my_other_vec <- c(22, 'orange', T)

my_other_vec
```

```{r}
my_third_vec <- c(T, F, 35)

my_third_vec
```

We can also missing values.

```{r}
my_fourth_vec <- c(1, 4, 5, NA)

my_fourth_vec

is.na(my_fourth_vec)
```

If I want to combine two vectors...

```{r}
a <- c(1, 2, 3)
b <- c(3, 5, 7)

c(a, b)
```

A brief example of matrices

```{r}
matrix(
  data = c(a, b),
  nrow = 2,
  byrow = T)
```

Sometimes I want special vectors, direct sequences of numbers. There are two ways to do this. If all I want is a integer sequence (made of doubles), then I use the "`<first number>:<last number>`".

```{r}
1:5
5:1
```

Other times, I need to count by something other than one, so I use `seq(from = <start>, to = <end>, by = <number to count by>)`

```{r}
seq(from = 1, to = 7, by = 1.3)
```

Hint: for brevity, I can leave off function parameter names, **as long as I enter them in order**

```{r}
seq(1, 7, by = 1.3)
```

If I add a constant to a vector, then they all go up by that constant.

```{r}
1:5 / 3
```

I can do math with equal-length sequences too.

```{r}
1:5 - seq(1, 4, by = .7)
```

But they **must** be equal lengths.

```{r, error=TRUE}
1:5 / 1:4
```

To access the elements of a vector, I put a number OR booleans in brackets `[]`.

```{r}
my_vec <- c('apple', 'orange', 'banana', 'pair')

my_vec[2]
```

```{r}
my_vec[2:4]
```

```{r}
my_vec[c(3, 2, 1, 4)]
```


I can also use bools.

```{r}
my_other_vec <- c(1, 4, 6, 7, 9, 3, 9)

my_other_vec < 5

my_other_vec[my_other_vec < 5]
```

I can also use functions that return values to access vectors, if I am creative...

```{r}
my_other_vec[max(my_other_vec) == my_other_vec]
```

R also has special vectors that are pre-loaded. The most commonly used are `letters` and `LETTERS`, which return the lower-case letters and uppercase letters of the English alphabet, respectively.


```{r}
vec <- c(1, 3, 4, 5, 3, 2, NA)

mean(vec, na.rm = T)
```


## Lists

Lists are a kind of vector that we make using the `list()` function, rather than the `c()` function. They have a few special properties.

## Lists

Lists are special vectors that can hold multiple types of elements.

They can even hold other vectors.

## Dataframes

### Construction

Dataframes are spreadsheets. Under the hood of R, they are just lists of vectors, where all the vectors are required to be the same length. To make one, you can call the `data.frame()` function and put your vectors inside.

```{r}
heights <- c(60, 65, 71, 72, 64)
sexes <- c('female', 'female', 'male', 'male', 'female')
shoes <- c('Adidas', 'Nike', 'Nike', 'Salvatore Ferragamo', 'Reebok')

df <- data.frame(height = heights, sex = sexes, shoes = shoes)

df
```

### Built-in dataframes

R has numerous built-in datasets that are ideal for demonstration purposes. We can get access to them using the `data()` command. This will load the data into our session, so we can then look at it.

```{r}
data('mtcars')

mtcars
```

Some datasets do not come in the form of a dataframe right away, but they can be converted into one using the `as.data.frame()` function.

```{r}
data(Seatbelts)

is.data.frame(Seatbelts)
```

```{r}
seatbelts_df <- as.data.frame(Seatbelts)

is.data.frame(seatbelts_df)
```


## Functions

A function is a piece of code that does work for you. It takes inputs and (usually) returns outputs. For example, the `sum()` function takes the mean of a numeric vector.

```{r}
my_vec <- c(3, 6, 2, 3)

sum(my_vec)
```

### Getting help

If I ever need to know something about a function, I can put a question mark in front of it (no `()`s) and run that line. That will bring up the help document for that function.

```{r}
?sum
```

### Function parameters

In addition to the data they take as input, most functions have additional *parameters* (sometimes called "arguments", but they mean the same thing). Looking at its help file, the `sum()` function has two parameters:

+ `...`, the numbers you want to sum
+ `na.rm = FALSE`, which tells `sum()` whether you want to remove ('rm') missing values ('na') before summing. 

Let's look at what happens when we try to `sum()` a vector with a missing value.

```{r}
my_vec <- c(5, NA, 2, 3) # should be 10

sum(my_vec)
```

R tells us the answer is missing (`NA`) because at least one of the vector elements is missing. This is to be conservative and to force you never to ignore missing values by accident. But what do we do if we really do want to sum all available values, ignoring the missing values.

Again, looking at the help file, we can see that the `na.rm` parameter of the function is followed by  `= FALSE`, under the Usage heading (see `sum(..., na.rm = FALSE)`). This tells us that the parameter `na.rm`, which tells `sum()` whether to remove missing values from the calulation, defaults to `FALSE`. 

To get `sum()` to ignore the missing values in our vector, we simply set `na.rm` to `TRUE` (or `T` for short).

```{r}
sum(my_vec, na.rm = T) # should be 10
```

## Packages

Packages are collections of functions that someone else put together for you. You can install them using the `install.packages()` function, with the name of your package inside the `()`.

```{r, eval=FALSE}
install.packages('ggplot2')
```

Once installed, use the `library()` function to load your package into your R session.

```{r}
library(ggplot2)
```

