---
title: "R Notebook"
output: html_notebook
---

# Functions

Functions allow you to automate boring tasks so that you don't have to keep doing the same thing over and over again. There are several advantages of this.

1. As stated above, it is faster for you.
2. When something changes (e.g., a reviewer wants you to exclude outliers), you only need to change your code in one place, rather than several.
3. You eliminate the risk of "copy-and-paste" mistakes.

This chapter covers the basics of function construction and use.

## Basic Example

```{r}
x <- 1:10
y <- 2:20
z <- 3:34


# Obscure reviewer request
x <- x/2
x <- sqrt(x)
x <- round(x)

y <- y/2
y <- sqrt(y)
y <- round(x)

z <- z/2
z <- sqrt(z)
z <- round(z)
```


## Anatomy of an R function

Making functions speeds up the process and makes it safer. To make a function, we need to use the `function()` command, assign it to a name, provide potential parameters, then give it code to execute.

```{r}
reviewer_transformation <- function(vals) {
  vals <- vals/2
  vals <- sqrt(vals)
  vals <- round(vals)
  
  vals 
}



# Obscure reviewer request
x <- 1:10
x <- x/2
x <- sqrt(x)
x <- round(x)

x
other_x <- 1:10
transformed_x <- reviewer_transformation(other_x)
```

```{r}
x
```


- Name
- Parameters/Arguments
    - Defaults
- Code
- Return values

$$
r = \frac{\text{cov}(x, y)}{\sigma_x\sigma_y}
$$


```{r}
# set.seed(314159)
x <- 1:10
y <- x*rpois(n = 10, lambda = 3)

w <- 1:20
u <- runif(n = 20)

cor(x, y)
cov(x, y)/(sd(x)*sd(y))

print('---')

correlation <- function(a, b) {
  cov(a, b)/(sd(a)*sd(b))
}

cor(w, u)
correlation(w, u)
```


## Base R functions

- Basic stat functions
- Random numbers!

```{r}
x <- rpois(100, 10)
y <- 2*x + rnorm(100)
df <- data.frame(x, y)

library(ggplot2)

ggplot(df, aes(x, y)) +
  geom_point() +
  geom_hline(yintercept = mean(y), color = 'red')
```

- Setting the random seed

- `paste()` and `paste0()` 

### Scoping

- What happens inside a function, stays inside a function
- Discussion of shadow variables

### Conditional execution

```{r}
set.seed(314159)
x <- 1:10
y <- x*rpois(n = 10, lambda = 3)

x[4] <- 1
y[5] <- 2

correlation <- function(a, b) {
  
  if(any(is.na(a)) & !any(is.na(b))){
    
    return('Only one missing data issue.')
  }
  else if(any(is.na(a)) & any(is.na(b))){
    
    return('You have missing data butthead!')
  } else {
    
    cov(a, b)/(sd(a)*sd(b))  
  }
}

cor(x, y)
correlation(a = x, b = y)
```


`if()`

`if()`/`else()`

`else if()`

`ifelse()`

### the `return()` command


--------------- Part II -----------------

## Pipes

Sometimes (often) it is useful to execute several functions in a row, using the answer from the last function as the input for the next function. But this can make your code redundant and a little unclear.

```{r}
reviewer_transformation <- function(vals) {
  vals <- vals/2
  vals <- sqrt(vals)
  vals <- round(vals)
  
  vals 
}

reviewer_transformation2 <- function(vals) {
  round(sqrt(vals/2))
}

reviewer_transformation_pipe <- function(vals) {
  vals %>%
    magrittr::divide_by(2) %>%
    sqrt() %>%
    round()
}

x <- 1:10
reviewer_transformation(x)
reviewer_transformation2(x)
reviewer_transformation_pipe(x)
```

```{r}
library(magrittr)

fun1 <- function(x){
  x *2
}

fun2 <- function(y) {
  y - 3
}

fun2(fun1(x))

x %>%
  fun1() %>%
  fun2()
```


To address this, we use pipes.

### Basic usage

The `%>%` operator (pronounced "pipe") takes whatever is on its left-hand-side (LHS) and applies it to the *first parameter* of the function listed on the right-hand-side. Install and load the `magrittr` package to use it.

```{r, eval=FALSE}
install.packages('magrittr')

library(magrittr)
```

### Sending the LHS to another parameter

But what do I do when I need the LHS output to go to another parameter in the RHS function? I use the `.` to remind the RHS function I want the LHS output to go to a different place.

<example using paste>

### Convenient magrittr functions


## More talk about the help files



## Functions from other packages

- I can import functions from another package by...
- What if I only want to import once?
- Namespace considerations

##
