---
title: "R Notebook"
output: html_notebook
---

# Reshaping data with `tidyr`

One of the most common problems in data management that lies just beyond most people's grasp is reshaping data. In this chapter, we will cover two basic issues.

1. Turning "long" data into "wide" data
2. Turning "wide" data into "long" data

If you are interested in a deeper understanding of how the shape of your data can improve the way you think about your analysis, [Hadley Wickam's paper on the subject](https://vita.had.co.nz/papers/tidy-data.pdf) is an excellent read.

## Turning long data into wide

Imagine you have an undergraduate research assistant who is trying to make life easy in a longitudinal study. In that study, participants return for a total of 3 times to retake the same test. You will eventually need each participant to have four columns - one for their name and three for each of their scores. However, the assistant has simply added a new row each time a participant returns, resulting in a dataset that looks like this:

```{r, message=FALSE, warning=FALSE}
library(tidyverse)

df <- expand.grid(id = c('Alice', 'Bob', 'Chantal'), time = c('a', 'b', 'c')) %>%
  arrange(id, time)
df$score <- sample(1:nrow(df))

df
```

To turn them into the format we want, we use the `spread()` function from the `tidyverse` sub-package "`tidyr`".

```{r}
spread_df <- df %>%
  spread(time, score)

spread_df
```

If I want to add the names of my key variable to my new column variables, I simply set the `sep` parameter (check the help files) to add the name of my variable to the new columns.

```{r}
df %>%
  spread(time, score, sep = '_')
```

## Turning wide data into long

To turn wide data back into long data, we use the `gather()` function from the same package. Note that `gather()` and `spread()` are exact opposites.

```{r}
long_df <- spread_df %>%
  gather(time, score)

long_df
```

Uh oh! We accidentally folded too much information into long format. We lost our id's into the time and score variables. This is because we forgot to tell `gather()` whether there were any columns that we *didn't* want to be involved in the process. In this case, we want our id's to stay our of the affair, so we add a `-` symbol in front of them.

```{r}
long_df <- spread_df %>%
  gather(time, score, -id)

long_df
```

Now, our data are returned to their original form.