---
title: "CSP 518 - Exercises: 1a - Intro to R"
author: "YOUR NAME HERE"
date: "Due: 2019-02-01 @ 11:59pm"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, eval=FALSE)
```

## Instructions

1. Make a .Rmd file to complete these exercises.

  - Title it "CSP 518 - Exercises: Intro to R" (in quotes)
  - Add your name(s) in the author slot (in quotes)
  - The output can be either be `pdf_document` or `html_document` (no quotes)

2. Turn in just your .Rmd file on Blackboard. Note, I will need to be able to `knit` it myself, so make sure everything runs the way you intend, before you turn it in.

3. For clarity, write "## Question <number>" befor each question to signal which one you are answering.

## Exercises

1. Create two string (character) variables called `a` and `b`, then concatenate them with `paste()`. The output should be to two strings together.

```{r}
a <- 'apples'
b <- 'bananas'

paste(a, b)
```

2. Create a vector called `my_vec` and store three numbers in it. Then, square all of those numbers. That is, if your vector is $1, 2, 3$, then the output should be $1, 4, 9$.

```{r}
my_vec <- 6:8
2*my_vec
```

3. Create two vectors of equal length and divide the first by the second.

```{r}
vec1 <- 4:6
vec2 <- 1:3

vec1/vec2
```

4. Print the first 13 letters of the alphabet, using brackets and a number sequence (hint, use the "`:`" trick)

```{r}
letters[1:13]
```

5. Create a vector of 10 numbers, ranging between 1 and 20. Then select all of them that are between 4 and 7.

```{r}
some_vec <- c(20 , 18 , 17 , 14, 10 , 4, 3, 5, 9, 2)

some_vec[(some_vec >= 4 & some_vec <= 7)]
```


6. With the same vector you used above, select every other number in the vector using `seq()` (starting with the first element).

```{r}
some_vec[seq(from = 1, to = length(some_vec), by = 2)]
```


7. Again, with the same vector from #5, select only the even numbers. (Hint: use modulo division, `%%`, to test whether the numbers have a remainder when divided by $2$).

```{r}
some_vec[some_vec %% 2 == 0]
```

8. Use the `mean()` function to find the mean of the following vector, ignoring the missing value.

```{r, echo=TRUE}
my_vec <- c(1, 4, 3, 5, 2, NA)
mean(my_vec, na.rm = T)
```
